# Descriptive statistics with Tidyverse

## Answers to Exercise Questions

### Exercise 1

Today is Monday. What day of the week will it be 9, 54, 306, and 8999
days from now?

Note: Create a character vector containing the days of the week and
repeat this vector 9000 times. Then, use indexing to find the desired
day. Hint: Write the days of the week in the character vector starting
from Tuesday.

```{r}
days <- c("Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday", "Monday")

# you complete...

days_rep <- rep(days,9000)

days_rep[9]
days_rep[54]
days_rep[306]
days_rep[8999]

# or::

days_rep[c(9,54,306,8999)]

```

### Exercise 2

Create a vector containing the numbers 1 to 100. Then, find the sum of
the numbers that are divisible by 3 or 5.

Tip: Use the %% operator to find the remainder of a division.

```{r}
# answer:
numbers <- 1:100

# find numbers that are divisible by 3:

numbers_3_5 <- numbers[numbers %% 3 == 0 | numbers %% 5 == 0]

sum(numbers_3_5)
```

### Exercise 3

You are taking measurements every 5 days throughout the year. Create a
number sequence that shows on which days you take measurements and
assign it to a variable named "measurement_days" The result should look
like this: 5, 10, 15, 20... 365.

```{r}
# answer:

days <- seq(5,365,5)
days[44]

```

## Practice questions for next week

***Q1:*** Create a vector containing 50 random numbers with a normal
(Gaussian) distribution, mean 20 and standard deviation 2. You can do
this with the `rnorm()` function. Then assigns the numbers to a variable
and use that variable as an argument to the `sample()` function to
randomly select 10 samples from that vector. Run `?rnorm()` `?sample()`
to see how the functions work and what arguments they take.

```{r, warning=FALSE, message=FALSE}

dist <- round(rnorm(n=50, mean=20, sd=2))

sample(dist, 10)
```

***Q2:*** Download and load "LearnBayes" package and take a look at the
first few columns of the data set called "studentdata".

***Answer the following questions:***

3.1. Remove rows that include NA observations.

```{r, warning=FALSE, message=FALSE}
#install.packages("LearnBayes")
library(LearnBayes)
df <- studentdata

dim(df)

sum(is.na(df))

df <- na.omit(df)

dim(df)

sum(is.na(df))

```

3.2. Get the number of female students.

```{r, warning=FALSE, message=FALSE}

# alt 1:
gender <- df$Gender
length(gender[gender == "female"])

# alt 2:
length(df$Gender[df$Gender == "female"])

# alt 3:
#install.packages("tidyverse")
library(tidyverse)
nrow(filter(df, Gender == "female"))

# Find percetange of female students
nrow(filter(df, Gender == "female")) / nrow(df)*100
```

3.3. Number of students who are taller than 180 cm (tip: the height is
given in inches. please first turn them to cm by multiplying the
observations with 2.54)

```{r, warning=FALSE, message=FALSE}
# You can use dplyr's mutate() function to change values in a dataset
library(dplyr)

df <- mutate(df, Height = Height * 2.54)

nrow(filter(df, Height > 180))
```

3.4. Plot the relationship between height and sex in a line graph.

```{r, warning=FALSE, message=FALSE}
plot(df$Gender, df$Height)
```

## Data Manipulation: In-class practice

Now, let's practice a bit more. Follow these steps:

We will now work with COVID-19 dataset. Let's import it to our R
session.

We can use dplyr's `read_csv()` package to import a spreadsheet in .csv
format.

*Source*: <https://github.com/sadettindemirel/Covid19-Turkey>

When working with data, it's important to create a workspace folder that
contains both your data and R script. You can either download the .csv
file and import it to R this way. Make sure that you set your working
directory to the correct folder by going to Session -\> Set Working
Directory -\> To Source File Location.

```{r, warning=FALSE, message=FALSE}

df_covid <- read_csv("http://kelesonur.github.io/compec-r/covid_sayi.csv")
head(df_covid)
```

1.  Create a histogram (of daily death toll) and time-series plot (x:
    time, y: daily death toll). You can use hist() and `boxplot()`
    functions respectively to do that. Add label names.

```{r, warning=FALSE, message=FALSE}
# Your turn:


```

2.  How many cases of COVID were detected and how many people died on
    July 20, 2020?

```{r, warning=FALSE, message=FALSE}
# Your turn:


```

3.  Can you show the total number of COVID deaths and patients in
    intensive care between 2020-03-11 and 2020-07-20 in the timeline
    graph? Tip: You can use indexing to filter the dates and save them
    as a new data frame.

```{r, warning=FALSE, message=FALSE}
# Your turn:


```

## Data Manipulation with Tidyverse

So far, we have examined data frames using Base R, that is, the
functions and operators native to R. However, today most data scientists
using R do not process data with these, but with the more modern
"Tidyverse" packages.

These packages make organizing data much easier and more practical
(Wickham, 2017).

Let's look at some important packages within Tidyverse:

1.  tibble: `as.tibble()`
2.  readr: `read_csv()`
3.  dplyr: `subset(), select(), filter(), summarize()` and more...
4.  magrittr: `%>%` operator
5.  ggplot2: `ggplot()`

```{r, warning=FALSE, message=FALSE}

library(tidyverse)
```

### Tibble

Tibble is essentially a more modern version of data frames in R. Here's
where Tibbles are superior to data frames: There is no need to use the
head() function. It automatically shows us the first 10 rows. There is
no need to use `nrow()` and `ncol()`. This information is included in
the table. tells us the type of vectors in each column (character,
integer, etc.)

```{r, warning=FALSE, message=FALSE}

df <- starwars
df <- as.tibble(df)
head(df)
```

### Data manipulation with dplyr

dplyr is the most practical package for editing data frames or tibbles.
Let's see what we can do with the functions included in it. For this we
will use a dataset from a package called `gapminder`:

```{r, echo=FALSE, warning=FALSE, message=FALSE}

library(gapminder)
library(kableExtra)

df <- gapminder

df <- head(df)
df %>% 
  kbl() %>%
  kable_styling()
```

Let's load the data:

```{r, warning=FALSE, message=FALSE}

library(gapminder)
  
df <- gapminder

```

#### `filter()` function

We can easily filter the data we want by using this function:

```{r, warning=FALSE, message=FALSE, results='hide'}
filter(df,country == "Turkey") 

filter(df, year > 2002) 

filter(df, year > 2002 & country == "Germany")

```

How can you filter for multiple countries? Any guesses?

#### `select()` function:

`select()` function allows us to select only the columns we want:

```{r, warning=FALSE, message=FALSE, results='hide'}

select(df, country, pop, year) # only tak

select(df, -continent) # kıta hariç bütün sütunları al

select(df, pop:gdpPercap) # nüfustan gdp'ye kadar bütün sütunları al
```

#### `rename()` function:

We can change the names of the columns we want with this. If we want
this to be saved, we need to overwrite the variable:

```{r, warning=FALSE, message=FALSE, results='hide'}
library(dplyr)
df <- rename(df, Country = country, CONTINENT = continent, YEAR = year, LifeExpectancy = lifeExp)

head(df)

# Extra: clean the column names
library(janitor)
df_cleaned <- clean_names(df)
```

#### `mutate()` function:

We can make changes to a tibble with the mutate() function.

For example, let's create a new column and write the population there as
one per million:

```{r, warning=FALSE, message=FALSE}

df_mutated <- mutate(df_cleaned, pop_million = pop / 1000000)

head(df_mutated)

```

#### `arrange()` function:

```{r, warning=FALSE, message=FALSE, results='hide'}

# sort increasingly

arrange(df_mutated, continent)

arrange(df_mutated, gdp_percap)

# sort in an descending order

arrange(df_mutated, desc(year))
```

### Pipe with the magrittr package

The pipe operator is represented by the symbol %\>%. This operator takes
the variable on its left and places it as the first argument of the
function on its right. This operator will be preferred instead of using
nested and long functions. This saves us from hard-to-read and nested
functions and helps us follow a linear order:

```{r, warning=FALSE, message=FALSE, results='hide'}

filter(df_mutated, country == "Turkey") # previous method

df_mutated %>% filter(country == "Turkey") # with pipe

select(filter(df_mutated, country == "Turkey" & year == "2007"), pop) # previous method

df_mutated %>% filter(country == "Turkey" & year == "2007") %>% select(pop) # pipe ile
```

### Visualization with the ggplot2 package

### Scatterplots

You can use `geom_point().`

```{r, warning=FALSE, message=FALSE}

library(ggplot2)

df_mutated %>% filter(year == 2007) %>%
ggplot(aes(x=gdp_percap, y=life_expectancy)) +
 geom_point() 

```

Or you can try using `geom_jitter()` or `geom_violin()`

```{r, warning=FALSE, message=FALSE}

df_mutated %>% filter(year == 2007) %>%
ggplot(aes(x=gdp_percap, y=life_expectancy)) +
 geom_violin() 

```

You can use text labels.

```{r, warning=FALSE, message=FALSE}

p1 <- df_mutated %>% filter(year == 2007) %>%
  ggplot(aes(x=gdp_percap, y=life_expectancy, label = country)) +
  geom_point() +
  geom_text()

p1 + ggtitle("Plot title") + xlab("GDP") + ylab("Life Exp")

# save it
# ggsave('plot1.png', width = 8, height = 6)

```

Using the color option:

```{r, warning=FALSE, message=FALSE}

df_mutated %>% filter(year == 2007) %>%
  ggplot(aes(x=gdp_percap, y=life_expectancy, label = country, color = continent)) +
  geom_point() +
  geom_text()

```

Add new facets:

```{r, warning=FALSE, message=FALSE}

 df_mutated %>% filter(year == 2007) %>%
  ggplot(aes(x=gdp_percap, y=life_expectancy, label = country, color = continent)) +
  geom_point() +
  geom_text() + 
  facet_wrap(~continent)
```

### Bar plots

```{r, warning=FALSE, message=FALSE}

df_mutated %>% filter(country %in% c("Turkey","Brazil","Thailand","Nigeria","New Zealand")) %>%
  ggplot(aes(x = reorder(continent,gdp_percap), y = gdp_percap, fill = continent)) +
  geom_bar(stat="identity")
```

## In-class Exercises

1.  What years of information are included in the Gapminder data? Tip:
    You can find this out with the `unique()` command.

```{r}
# Your answer:


```

2.  What are the mean and median of life years in 1962 and 2002? Tip:
    You can do this with the `filter()` and `select()` functions.

```{r}
# Your answer:


```

3.  Can you create a scatterplot of income (x-axis) and years of life
    expectancy (y-axis) for the European continent in 1960, 1980 and
    2000 using `ggplot()` and `facet_wrap()`? Country names must appear
    as text. Also color by continent. Tip: after filtering the data by
    these years you will need to call `facet_wrap(~year)`.

```{r}
# Your answer:


```

4.  Can you create a line plot and show Turkey's population growth? Can
    you name the plot and axes? Tip: you will need to use the
    `geom_line()` function to create the line chart.

```{r}
# Your answer:


```
